Wait-free Queues with Polylogarithmic Step Complexity

### 2022-09-01 thesis/thesis.tex
new code for indexing super

\Statex
\Statex $\diamondsuit$ \tt{\sl{Structures}}

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
\item \tt{\sl{BlockList}}
\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent.}
\end{itemize}

@@

  \item \tt{\sl{int} group}
  \textsf{: the value read from \tt{num\sub{propagated}} when appending this block to the node.}
\end{itemize}

@@

\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{<new, np\sub{left}, np\sub{right}>= \Call{CreateBlock}{h}} \label{invokeCreateBlock} \cmt{\tt{np\sub{left}, np\sub{right}} are the values read from the children's \tt{num\sub{propagated}} field.}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} \cmt{The block contains nothing.}
\ElsIf{\tt{blocks.tryAppend(new, h)}} \label{cas}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{\Call{CAS}{dir.super[np\sub{dir}], null, h}} \cmt{Write would work too.} \label{setSuper}
\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}} \label{incNP}
\EndFor
\State \tt{\Call{CAS}{\head, h, h+1}} \label{incrementHead1}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{\head, h, h+1}} \cmt{Even if another process wins, help to increase the \tt{\head}. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

@@

\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} $\geq$ \tt{i}}
\Function{int}{BSearch}{\sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\cmt{Creates a block to be inserted as $i$th \tt{block} in \tt{blocks}. Returns the created block as well as values read from each child's \tt{num\sub{propagated}} field. These values are used for incrementing the children's \tt{num\sub{propagated}} field if the block was appended to \tt{blocks} successfully.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\State \tt{newBlock.group= num\sub{propagated}}\label{setGroup}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.\head-1} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{newBlock.end\sub{dir}= index\sub{last}} \label{endDefLine}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{np\sub{dir}= dir.num\sub{propagated}} \label{setNP}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\EndFor
%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
%\State \tt{b.sum= n.blocks[i-1].sum + b.num}

@@

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_{this,b,i}=D_{root,x,y}$.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>} \label{indexBaseCase}
\Else
\State \tt{dir= (parent.left==n)? left: right} \cmt{check if this node is a left or a right child}
\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, super[blocks[b].group]-p, super[blocks[b].group]+p}} \label{computeSuper}
\Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}

@@

\end{proof}
\begin{lemma}[DSearch Analysis]\label{dsearchTime}
  Assume $\nf{root.blocks[end].sum\sub{enq}} \geq \nf{e}$ and $E_e(root)$'s \nf{element} is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}, then  \nf{DSearch(e, end)} takes $\Theta(\log$\nf{root.blocks[b].\size+ root.blocks[end].\size}$)$ steps.
\end{lemma}
\begin{proof}
First we show $\nf{end} - \nf{b} \leq 2 \times \big(\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size}+1 \big)$. From line \ref{addOP}, we know that \nf{num} field of the every block in the tree is greater than 0. So, each block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue} or at least one \nf{Dequeue}. Suppose there were more than \nf{root.blocks[b].\size}\nf{Dequeue}s in \nf{root.blocks[b+1..end-1]}. Then the element in the queue which is the response to the \nf{Dequeue()} would become dequeued at some point after \nf{blocks[b]}'s last operations and before \nf{root.blocks[end]}'s first operation. Which means the response to to a \nf{Dequeue} in \nf{root.blocks[end]} could not be in $E(n,b)$. Furthermore since the size of the queue would become \nf{root.blocks[end].\size} after the operations of \nf{root.blocks[end]}, there cannot be more than \nf{root.blocks[b].\size + root.blocks[end-1].\size} \nf{Enqueue}s in \nf{root.blocks[b+1..end-1]}. Because there can be at most \nf{root.blocks[b].\size}\nf{Dequeue}s and the final size of the queue is \nf{root.blocks[end-1].\size}. Overall there can be at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} operations in \nf{root.blocks[b+1..end-1]} and since each block size is $\geq 1$ thus there are at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]}. So $\nf{end-b}\leq 2 \times\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size}+1$. See Figure \ref{fig::doubling}.

>>>>

%\Statex $\diamondsuit$ \tt{\sl{Structures}}

\Statex $\blacktriangleright$ \tt{\sl{Node}}
\begin{itemize}
\item \tt{\sl{*Node} left, right, parent} \textsf{: initialized  when creating the tree.}
\item \tt{\sl{BlockList}}
\item \tt{\sl{int} \head= 1}\textsf{: \#\tt{block}s in \tt{blocks}. \tt{blocks[0]} is a block with all integer fields equal to zero.}
%\item \tt{\sl{int} num\sub{propagated}= 0}\textsf{} \textsf{: \# groups of blocks that have been propagated from the node to its parent.}
\end{itemize}

@@

  \item \tt{\sl{int} super}
  \textsf{: approximate index of the superblock, read from \tt{parent.head} when appending the block to the node}
\end{itemize}

@@

\Function{boolean}{Refresh()}{}
\State \tt{h= \head} \label{readHead}
\State \tt{hp= parent.\head} \label{readParentHead1}
\State \tt{CAS(blocks[h-1].super, null, hp)}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{h\sub{dir}= dir.\head} \label{readChildHead}
\State \tt{CAS(dir.blocks[h\sub{dir}], null, h)}
\EndFor
%\If{\tt{n.blocks[h]!=null?}} \tt{h+=1}\EndIf
\State \tt{new= \Call{CreateBlock}{h}} \label{invokeCreateBlock}
\If{\tt{new.num==0}} \Return{\tt{true}} \label{addOP} \cmt{Created block contains nothing.}
\ElsIf{\tt{blocks.tryAppend(new, h)}} \label{cas}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}} \label{okcas}
\State \tt{hp= parent.\head} \label{readParentHead2}
\State \tt{\Call{CAS}{new.super, null, hp}} \label{setSuper}
%\State \tt{\Call{CAS}{dir.num\sub{propagated}, np\sub{dir}, np\sub{dir}+1}} \label{incNP}
\EndFor
\State \tt{\Call{CAS}{\head, h, h+1}} \label{incrementHead1}
\State \Return{\tt{true}}
\Else
\State \tt{\Call{CAS}{\head, h, h+1}} \cmt{Even if another process wins, help to increase the \tt{\head}. The winner might have fallen sleep before increasing \tt{\head}.} \label{incrementHead2}
\State \Return{ \tt{false}}
\EndIf
\EndFunction{Refresh}

\pagebreak
\Statex $\leadsto$ \textsf{Precondition: \tt{blocks[start..end]} contains a block with field \tt{f} $\geq$ \tt{i}}
\Function{int}{BSearch}{\sl{field} f, \sl{int} i, \sl{int} start, \sl{int} end}

@@

\Function{<Block, int, int>}{CreateBlock}{\sl{int} i} 
\cmt{Creates and returns the block to be inserted as $i$th \tt{block} in \tt{blocks}.}
\State \tt{block newBlock= \Call{new}{\sl{block}}}
\ForEach{\tt{dir} {\keywordfont{in}} \tt{\{left, right\}}}
\State \tt{index\sub{last}= dir.\head-1} \label{lastLine}
\State \tt{index\sub{prev}= blocks[i-1].end\sub{dir}} \label{prevLine}
\State \tt{newBlock.end\sub{dir}= index\sub{last}} \label{endDefLine}
\State \tt{block\sub{last}= dir.blocks[index\sub{last}]}
\State \tt{block\sub{prev}= dir.blocks[index\sub{prev}]}
\State \cmt{\tt{newBlock} includes \tt{dir.blocks[index\sub{prev}+1..index\sub{last}]}.}
\State \tt{newBlock.sum\sub{enq-dir}= blocks[i-1].sum\sub{enq-dir} + block\sub{last}.sum\sub{enq} - block\sub{prev}.sum\sub{enq}}
\State \tt{newBlock.sum\sub{deq-dir}= blocks[i-1].sum\sub{deq-dir} + block\sub{last}.sum\sub{deq} - block\sub{prev}.sum\sub{deq}}
\EndFor
%\State \tt{b.num\sub{enq}= b.num\sub{enq-left} + b.num\sub{enq-right}}
%\State \tt{b.num\sub{deq}= b.num\sub{deq-left} + b.num\sub{deq-right}}
%\State \tt{b.num= b.num\sub{enq} + b.num\sub{deq}}
%\State \tt{b.sum= n.blocks[i-1].sum + b.num}

@@

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
\Function{<int, int>}{IndexDeq}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_{this,b,i}=D_{root,x,y}$.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>} \label{indexBaseCase}
\Else
\State \tt{dir= (parent.left==n)? left: right} \cmt{check if this node is a left or a right child}
\State \tt{superBlock= parent.\Call{BSearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, blocks[b].super-2, blocks[b].super+2}} \label{computeSuper}
\Statex\cmt{superblock's group has at most $p$ difference with the value stored in \tt{super[]}.}

@@

\end{proof}
\begin{lemma}[DSearch Analysis]\label{dsearchTime}
  Assume $\nf{root.blocks[end].sum\sub{enq}} \geq \nf{e}$ and $E_e(root)$'s \nf{element} is the response to some \nf{Dequeue()} operation in \nf{root.blocks[end]}, then  \nf{DSearch(e, end)} takes $\Theta(\log$\nf{root.blocks[b].\size+ root.blocks[end].\size}$)$ steps.
\end{lemma}
\begin{proof}
First we show $\nf{end} - \nf{b} \leq 2 \times \big(\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size}+1 \big)$. From line \ref{addOP}, we know that \nf{num} field of the every block in the tree is greater than 0. So, each block in \nf{root.blocks[b..end]} contains at least one \nf{Enqueue} or at least one \nf{Dequeue}. Suppose there were more than \nf{root.blocks[b].\size}\nf{Dequeue}s in \nf{root.blocks[b+1..end-1]}. Then the element in the queue which is the response to the \nf{Dequeue()} would become dequeued at some point after \nf{blocks[b]}'s last operations and before \nf{root.blocks[end]}'s first operation. Which means the response to to a \nf{Dequeue} in \nf{root.blocks[end]} could not be in $E(n,b)$. Furthermore since the size of the queue would become \nf{root.blocks[end].\size} after the operations of \nf{root.blocks[end]}, there cannot be more than \nf{root.blocks[b].\size + root.blocks[end-1].\size} \nf{Enqueue}s in \nf{root.blocks[b+1..end-1]}., because there can be at most \nf{root.blocks[b].\size}\nf{Dequeue}s and the final size of the queue is \nf{root.blocks[end-1].\size}. Overall there can be at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} operations in \nf{root.blocks[b+1..end-1]} and since each block size is $\geq 1$ thus there are at most $2 \times$\nf{root.blocks[b].\size+ root.blocks[end].\size} blocks in between \nf{root.blocks[b]} and \nf{root.blocks[end]}. So $\nf{end-b}\leq 2 \times\nf{root.blocks[b].\size}+\nf{root.blocks[end].\size}+1$. See Figure \ref{fig::doubling}.
